A few stickleback suffice to transport adaptive alleles to new lakes

### 2018-12-09 Stickleback_Paper.tex
lots of words about prob of fixation that we do not need now, grr

@@

An alternative explanation is that at lower rates of gene flow,
freshwater alleles present as standing variation in the marine habitat
are more tightly linked to marine alleles,
and so adaptation in the new lakes must wait for recombination.
This might be expected since freshwater alleles present in the ocean,
to remain present at migration-selection balance at low migration rates,
must be ``masked'' by nearby compensatory alleles.
Reversing this ``masking'' then slows the process of adaptation that uses this genetic variation.
Said another way,
higher gene flow from freshwater to the ocean
maintains relatively intact freshwater haplotypes
that can be more easily rebuilt.
To confirm this, we calculated for each individual in the initial generation of the new lakes
the total contribution to trait values of each of the ten genomic regions 
that containing trait-affecting mutations.
The recombination rate within each of these ten regions is $10^{-2}$,
so these regions act initially more-or-less like single loci.
Even though the trait is additive across loci,
the presence of sufficiently many distinct alleles to attain, in principle,
the optimal freshwater trait value
does not ensure adaptation will actually occur with these alleles.
On the contrary, many of these will likely be lost to drift.
To get a rough idea of whether the genetic variation present in the initial population
is sufficient for the population to adapt,
we translated the total effect size $x$ of each region 
to a probability of fixation
(by $p(x) \approx 2x/45$),
and then calculated $\sum_i x_i p(x_i)$.
If a region moves the trait an average of $x$ units towards the freshwater optimum
(averaging over dominance relationships),
Given the probability that an ``effect region'' fixes at a given trait value, 
$2 \times x / 45$ where $x$ is the effect size of a haplotype,
we were able to determine expected total effect sizes of the haplotypes that fix (detail about computation in methods).
Below, we see the mean of these expected total effect sizes across all lakes.

@@

Suppose a new allele enters a lake, either by migration or mutation. If, when it is rare but present in $n$ copies, it has fitness advantage $s$ -- i.e., the expected number of copies in the next generation is $(1+s)n$ --then the probability that it escapes demographic stochasticity to become common in the population is approximately $2s$ \citep{fisher,prob_fixation}. If the current population all differed from the optimum trait by $z$, and the allele has effect size $-u$ in heterozygotes, then the fitness advantage of the allele would be $s(u) = \exp(-\beta((z - u)^2) / \exp( - \beta z^2)) \approx 2 \beta z u$, where in our parameterization, $\beta = 1 / 450$. This tells us two things: (1) the rate of adaptation decreases as the population approaches the optimum, and (2) larger mutations (in the right direction) are more likely to fix.

>>>>

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}

@@

An alternative explanation is that at lower rates of gene flow,
freshwater alleles present as standing variation in the marine habitat
are more tightly linked to marine alleles,
and so adaptation in the new lakes must wait for recombination.
This might be expected since freshwater alleles present in the ocean,
to remain present at migration-selection balance at low migration rates,
must be ``masked'' by nearby compensatory alleles.
Reversing this ``masking'' then slows the process of adaptation that uses this genetic variation.
Said another way,
higher gene flow from freshwater to the ocean
maintains relatively intact freshwater haplotypes
that can be more easily rebuilt.
Recall that trait-affecting mutations only occur in relatively small regions of $10^5$ loci
in which recombination occurs at rate $10^{-2}$.
Even if there is a sufficient amount of variation in the initial population of a lake
to shift the trait from the marine optimum ($+10$) to the freshwater optimum ($-10$),
these may be tightly linked to alleles that counteract their effect.
For example, suppose there are 10 variants segregating at low frequency with effect size $-1$ each,
but each is paired with a compensatory allele with effect size $+1$.
Each local haplotype is therefore neutral, 
and so likely to be at higher frequency in the marine population.

Suppose that the alleles present in the initial population of a new lake
have effect sizes $(x_1, \ldots, x_n)$,
and that the probability of fixation of a single allele of effect size $x$ is $p(x)$.
Let $X_i$ be the average per capita effect of the $i^\text{th}$ allele after fixation,
so that $X_i = 0$ if the allele is lost,
and $X_i = 2x_i$ if the allele fixes (ignoring dominance).
Therefore, the expected contribution to the final trait value of allele $i$
is $\E[X_i] = 2 x_i p(x_i)$.
The trait value after the initial round of adaptation is $\sum_i X_i$,
and so the expected trait value is $2 \sum_i x_i p(x_i)$.
Since there are many alleles that contribute more-or-less additively to the trait,
the probability of fixation of a particular allele is not easy to find,
but we can roughly approximate it as follows:
if the current population trait mean $z$,
then an allele that moves the trait a distance $x$ closer to the optimum
has a fitness advantage in heterozygotes of 
$\exp(-\beta((z - x)^2) / \exp( - \beta z^2)) \approx 1 + 2 \beta z x$, 
where in our parameterization, $\beta = 1 / 450$.
Since the probability of fixation of a single allele with relative advantage $s$
is approximately $2s$ \citep{lambert2006probability,haldane1927mathematical},
then with $z=10$ we can use $p(x) \approx 4 x / 45$.
Putting these considerations together,
a rough estimate for the total shift in trait value resulting from selection
on standing variation in the initial population is
$R = \sum_i 8 x_i / 45$,
where the sum is over all trait-affecting alleles.
This approximation should be good if each trait-affecting allele is rare (present in only a few copies),
effects are additive, and most are unlinked.

We computed $R$ to confirm our suspicion that linkage slowed down adaptation
by showing that there is sufficient allelic variation in the initial populations
to complete adaptation across both migration rates (0.1 and 1 migrants per lake per generations),
but not if the haplotypes of each genomic region are assumed to not recombine.

We calculated for each individual in the initial generation of the new lakes
the total contribution to trait values of each of the ten genomic regions 
that containing trait-affecting mutations.
The recombination rate within each of these ten regions is $10^{-2}$,
so these regions act initially more-or-less like single loci.
Even though the trait is additive across loci,
the presence of sufficiently many distinct alleles to attain, in principle,
the optimal freshwater trait value
does not ensure adaptation will actually occur with these alleles.
On the contrary, many of these will likely be lost to drift.

To get a rough idea of whether the genetic variation present in the initial population
is sufficient for the population to adapt,
we translated the total effect size $x$ of each region 
to a probability of fixation
(as described below $p(x) \approx 2x/45$),
and then calculated $\sum_i x_i p(x_i)$.
If a region moves the trait an average of $x$ units towards the freshwater optimum
(averaging over dominance relationships),
Given the probability that an ``effect region'' fixes at a given trait value, 
$2 \times x / 45$ where $x$ is the effect size of a haplotype,
we were able to determine expected total effect sizes of the haplotypes that fix (detail about computation in methods).
Below, we see the mean of these expected total effect sizes across all lakes.

@@

Suppose a new allele enters a lake, either by migration or mutation. 
If, when it is rare but present in $n$ copies, it has fitness advantage $s$ 
-- i.e., the expected number of copies in the next generation is $(1+s)n$ 
-- then the probability that it escapes demographic stochasticity to become common in the population 
is approximately $2s$ \citep{fisher,prob_fixation}. 
If the current population all differed from the optimum trait by $z$, 
and the allele has effect size $-u$ in heterozygotes, 
then the fitness advantage of the allele would be $s(u) = \exp(-\beta((z - u)^2) / \exp( - \beta z^2)) \approx 2 \beta z u$, where in our parameterization, $\beta = 1 / 450$. This tells us two things: (1) the rate of adaptation decreases as the population approaches the optimum, and (2) larger mutations (in the right direction) are more likely to fix.
